Dynamics of Colombo's Top: Generating Exoplanet Obliquities from Planet-Disc Interactions

### 2020-04-23 writeups/Su_disk_disp.tex
check in new explanation for nonad sdi > 0 case

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
        caption,
        subcaption,
    }
    \usepackage[T1]{fontenc}
    \captionsetup{compatibility=false}

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\pdil}[2]{\partial#1 / \partial#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\bm{#1}}}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

% for i in ../initial/99_misc/2_3vec.png ../initial/99_misc/2_cs_locs.png ../initial/0_eta/1contours_flip.png ../initial/99_misc/1_areas.png ../initial/2_toy2/3_ensemble_05_35.png ../initial/2_toy2/3_ensemble_10_35.png ../initial/2_toy2/3_ensemble_20_35.png ../initial/2_toy2/3testo21.png ../initial/2_toy2/3testo21_subplots.png ../initial/2_toy2/3testo23.png ../initial/2_toy2/3testo23_subplots.png ../initial/2_toy2/3testo31.png ../initial/2_toy2/3testo31_subplots.png ../initial/2_toy2/3testo321.png ../initial/2_toy2/3testo321_subplots.png ../initial/2_toy2/3_ensemble_05_25.png ../initial/2_toy2/3_ensemble_05_15.png ../initial/2_toy2/3testo_nonad.png ../initial/2_toy2/3testo_nonad_subplots.png ../initial/2_toy2/3_ensemble_05_05.png ../initial/2_toy2/3scan.png ../initial/2_toy2/3scan_20.png ../initial/99_misc/2_lambdas.png; do cp $i plots_diskdisp; done

@@

Spin states satisfying $\p{d\uv{s}/d\uv{\tau}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs) \citep{colombo1966,peale1969}. They require that
$\uv{s}$, $\uv{l}$, and $\uv{l}_{\rm d}$ be coplanar. There are either two or
four CSs, depending on the value of $\eta$. They are specified by the obliquity
$\theta$ and the precessional phase of $\uv{s}$ around $\uv{l}$, denoted by
$\phi$. Following the standard convention and nomenclature (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and
$\theta < 0$, corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite
sides of $\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to
$\uv{s}$ and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$. The CS
obliquity satisfies
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.
\end{equation}
When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.

@@

In practice, as $\eta$ evolves, the system only crosses the separatrix once or
twice, while it spends many orbits inside one of the three zones and far from
the separatrix. Thus, a \emph{weak adiabaticity criterion} is that the variation
timescale of $\eta$ be slower than all circulation/libration timescales near the
equilibria/fixed points. If this criterion is satisfied, then the system will
evolve adiabatically for most of its evolution save one or two separatrix
crossings.

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it has the smallest characteristic frequency in the
system. The weak adiabaticity criterion can be recast as requiring that $\eta$
vary more slowly than this libration frequency \citep[see][]{ward2004II}, i.e.\
$t_{\rm d}^{-1} \lesssim \omega_{\rm lib} / 2\pi$. In terms of the dimensionless
parameter $\epsilon$
[Eq.~\eqref{eq:eps_def}], this adiabaticity condition is
\begin{equation}
    \epsilon \lesssim \epsilon_{\rm c} \equiv
        \frac{1}{2\pi}\sqrt{\eta\sin I \sin \theta_2
            \p{1 + \eta \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{equation}
where $\theta_2$ is the obliquity at CS2. This formula differs from that given
in \citet{millholland_disk}, where the $\csc^3\!\theta_2$ term is neglected and
the square root is missing\footnote{The missing $\csc^3 \theta_2$ term can be
traced to a $\theta \gg I$ approximation made in Eq.~(3) of \citet{ward2004II}.
Since $\theta_2 \approx I$ for $\eta \gg 1$ (Fig.~\ref{fig:cs_locs}), this
approximation is not always valid.}.

For $I = 5^\circ$, $\eta \sim \eta_{\rm c}$ Eq.~\eqref{eq:ad_constr} gives
$\epsilon_{\rm c} = 0.085$. Since our criterion is only a weak condition for
adiabaticity, we use $\epsilon = 3 \times 10^{-4}$ in our ``adiabatic''
calculations below. We explore the consequences of nonadiabatic evolution in
Section~\ref{s:nonad}.

@@

The above calculation applies for a specific initial $\theta_{\rm i}$, but, as
discussed at the beginning of Section~\ref{ss:ad_ensemble}, the initial spin
orientation is more appropriately described by $\theta_{\rm sd, i}$ since
$\eta_i\gg 1$. The correct way to predict the final obliquity for a given
$\theta_{\rm sd, i}$ using Eq.~\eqref{eq:thetaf}
is somewhat indirect but yields good agreement with numerical results.

First consider the case with $\theta_{\rm sd, i} = 0$. This corresponds to a
well-defined initial obliquity $\theta_i = I$ (more precisely, the initial
condition is CS2). Thus
\begin{align}
    \sin\theta_{\rm f} &\simeq \sin I \abs{1 + e^{-i\varphi_0}
            \sqrt{\frac{2\pi \cos I}{\epsilon}}},\nonumber\\
        &\approx \sin I \sqrt{\frac{2\pi \cos I}{\epsilon}},
            \label{eq:theta_2f}
\end{align}
where the second equality assumes $\sqrt{2\pi/\epsilon}\gg 1$.
Fig.~\ref{fig:nonad_3_scan} shows the final obliquity as function of $\epsilon$
for $\theta_{\rm sd,i} = 0$ and $I = 5^\circ$. We see that the agreement between
the numerical results and Eq.~\eqref{eq:theta_2f} is excellent. For $\epsilon
\ll \epsilon_{\rm c}$, we find $\theta_{\rm f} \simeq 90^\circ$, in agreement
with the result of adiabatic evolution (see Fig.~\ref{fig:ad_ensemble}).

For $\theta_{\rm sd, i}\neq 0$, the set of possible initial conditions forms a
ring of angular radius $\theta_{\rm sd, i}$ about $\uv{l}_{\rm d}$ (or CS2). We
find that this set of initial conditions approximately evolves under
nonadiabatic evolution to a ring of angular width $\theta_{\rm sd, i}$ about the
$\theta_{\rm sd, i} = 0$ result given by Eq.~\eqref{eq:theta_2f}. Intuitively,
because of the rapid precession of $\uv{l}$ around $\uv{l}_d$ at $\eta = \eta_i
\gg 1$, the initial $\theta_i$ varies rapidly but is centered around $\theta_i
\simeq I$. Thus the final obliquity spans in the range
\begin{equation}
    \abs{\sqrt{\frac{2\pi \cos I}{\epsilon}}\sin I
            - \sin\theta_{\rm sd, i}}
        \lesssim \sin\theta_{\rm f}
        \lesssim
            \sqrt{\frac{2\pi \cos I}{\epsilon}}\sin I + \sin \theta_{\rm sd, i}
    .\label{eq:nonad_q_f_dist}
\end{equation}
Fig.~\ref{fig:nonad_3_ensemble} shows the numerical result of $\theta_f$ vs
$\theta_{\rm sd,i}$ for $I =5^\circ$ and $\epsilon = 0.3$. We see that
Eq.~\eqref{eq:nonad_q_f_dist} provides good lower and upper bounds of the final
obliquity for $\theta_{\rm sd,i} \lesssim 45^\circ$.

>>>>

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
        caption,
        subcaption,
        xcolor,
    }
    \usepackage[T1]{fontenc}
    \captionsetup{compatibility=false}

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\rdil}[2]{\mathrm{d}#1 / \mathrm{d}#2}
    \newcommand*{\pdil}[2]{\partial#1 / \partial#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\bm{#1}}}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

    \colorlet{Corr}{red}

@@

% for i in ../initial/99_misc/2_nonad_rot.png ../initial/99_misc/2_3vec.png ../initial/99_misc/2_cs_locs.png ../initial/0_eta/1contours_flip.png ../initial/99_misc/1_areas.png ../initial/2_toy2/3_ensemble_05_35.png ../initial/2_toy2/3_ensemble_10_35.png ../initial/2_toy2/3_ensemble_20_35.png ../initial/2_toy2/3testo21.png ../initial/2_toy2/3testo21_subplots.png ../initial/2_toy2/3testo23.png ../initial/2_toy2/3testo23_subplots.png ../initial/2_toy2/3testo31.png ../initial/2_toy2/3testo31_subplots.png ../initial/2_toy2/3testo321.png ../initial/2_toy2/3testo321_subplots.png ../initial/2_toy2/3_ensemble_05_25.png ../initial/2_toy2/3_ensemble_05_15.png ../initial/2_toy2/3testo_nonad.png ../initial/2_toy2/3testo_nonad_subplots.png ../initial/2_toy2/3_ensemble_05_05.png ../initial/2_toy2/3scan.png ../initial/2_toy2/3scan_20.png ../initial/99_misc/2_lambdas.png; do cp $i plots_diskdisp; done

@@

Spin states satisfying $\p{d\uv{s}/d\uv{\tau}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs) \citep{colombo1966,peale1969}. They require that
$\uv{s}$, $\uv{l}$, and $\uv{l}_{\rm d}$ be coplanar. There are either two or
four CSs, depending on the value of $\eta$. They are specified by the obliquity
$\theta$ and the precessional phase of $\uv{s}$ around $\uv{l}$, denoted by
$\phi$. Following the standard convention and nomenclature (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and
$\theta < 0$, corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite
sides of $\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to
$\uv{s}$ and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$. The CS
obliquity satisfies
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.\label{eq:cs_zero}
\end{equation}
When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.

@@

In practice, as $\eta$ evolves, the system only crosses the separatrix once or
twice, while it spends many orbits inside one of the three zones and far from
the separatrix. Thus, a \emph{weak adiabaticity criterion} is
that\textcolor{Corr}{, for all equilibria/fixed points, the local
circulation/libration timescales are much shorter than motion of the
equilibrium}. If this criterion is satisfied, then the system will evolve
adiabatically for most of its evolution save one or two separatrix crossings.

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it has the smallest characteristic frequency in the
system. The weak adiabaticity criterion \textcolor{Corr}{is equivalent to
requiring that, for all times other than separatrix crossing, the obliquity of
CS2 ($\theta_2$) evolve over longer timescales than the local libration period
about CS2, or
\begin{equation}
    \rd{\theta_2(t)}{t} \lesssim \frac{\omega_{\rm lib}(t)}{2\pi},\label{eq:ad_base}
\end{equation}
where
\begin{equation}
    \omega_{\rm lib}(t) = \sqrt{\eta(t)\sin I \sin \theta_2(t)
            \p{1 + \eta \sin I \csc^3 \theta_2(t)}},\label{eq:w_lib}
\end{equation}
is the libration frequency about CS2 at time $t$ (Appendix~\ref{ss:eigens}).}
This formula differs from that given in \citet{millholland_disk}, where the
$\csc^3\!\theta_2$ term is neglected and the square root is missing\footnote{The
missing $\csc^3\!\theta_2$ term can be traced to a $\theta \gg I$ approximation
made in Eq.~(3) of \citet{ward2004II}. Since $\theta_2 \sim I$ for $\eta \sim 1$
(Fig.~\ref{fig:cs_locs}), this approximation is not always valid.}.
\textcolor{Corr}{Differentiating Eq.~\eqref{eq:cs_zero} gives
\begin{equation}
    \rd{\theta_2}{t} = \epsilon \frac{\eta \sin\p{\theta_2(t) - I}}{
        \cos\p{2\theta_2(t)} - \eta \cos\p{\theta_2(t) - I}},\label{eq:dq2dt}
\end{equation}
where $\epsilon = -\rdil{(\ln \eta)}{t}$ [Eq.~\eqref{eq:eps_def}]. For a given
$\epsilon$, Eqs.~(\ref{eq:ad_base}--\ref{eq:dq2dt}) will be satisfied for all
times if they are satisfied near $\eta \simeq 1$, where $\dot{\theta}_2 \simeq
\epsilon$. Thus, weak adiabiticity requires
\begin{equation}
    \epsilon \lesssim \epsilon_{\rm c},\label{eq:ad_constr}
\end{equation}
where
\begin{equation}
    \epsilon_{\rm c} \equiv \p{\frac{\omega_{\rm lib}}{t}}_{\eta = 1} =
        \frac{1}{2\pi}\sqrt{\sin I \sin \theta_{\rm 2c}
            \p{1 + \sin I \csc^3 \theta_{\rm 2c}}},
\end{equation}
where $\theta_{\rm 2c}$ is the location of CS2 for $\eta = 1$.
% theta_{2c} = 31^\circ$
} For $I = 5^\circ$, we find $\epsilon_{\rm c} = 0.0431$. Since our criterion is
only a weak condition for adiabaticity, we use $\epsilon = 3 \times 10^{-4}$ in
our ``adiabatic'' calculations below. We explore the consequences of
nonadiabatic evolution in Section~\ref{s:nonad}.

@@

The above calculation applies for a specific initial $\theta_{\rm i}$, but, as
discussed at the beginning of Section~\ref{ss:ad_ensemble}, the initial spin
orientation is more appropriately described by $\theta_{\rm sd, i}$ since
$\eta_i\gg 1$. The correct way to predict the final obliquity for a given
$\theta_{\rm sd, i}$ using Eq.~\eqref{eq:thetaf}
is somewhat \textcolor{Corr}{subtle} but yields good agreement with numerical
results.

First consider the case with $\theta_{\rm sd, i} = 0$. This corresponds to a
well-defined initial obliquity $\theta_i = I$ (more precisely, the initial
condition is CS2). \textcolor{Corr}{The final obliquity in this case, denoted
$\theta_{\rm 0f}$, is given by}
\begin{align}
    \sin\theta_{\rm 0f} &\simeq \sin I \abs{1 + e^{-i\varphi_0}
            \sqrt{\frac{2\pi \cos I}{\epsilon}}},\nonumber\\
        &\approx \sin I \sqrt{\frac{2\pi \cos I}{\epsilon}},
            \label{eq:theta_2f}
\end{align}
where the second equality assumes $\sqrt{2\pi/\epsilon}\gg 1$.
Fig.~\ref{fig:nonad_3_scan} shows the final obliquity as function of $\epsilon$
for $\theta_{\rm sd,i} = 0$ and $I = 5^\circ$. We see that the agreement between
the numerical results and Eq.~\eqref{eq:theta_2f} is excellent. For $\epsilon
\ll \epsilon_{\rm c}$, we find $\theta_{\rm f} \simeq 90^\circ$, in agreement
with the result of adiabatic evolution (see Fig.~\ref{fig:ad_ensemble}).

\begin{figure}
    \centering
    \includegraphics[width=0.6\columnwidth]{plots_diskdisp/2_nonad_rot.png}
    \caption{\textcolor{Corr}{Schematic depicting the calculation of
    $\theta_{\rm f}$ for nonadiabatic evolution when $\theta_{\rm sd, i} > 0$,
    where $\theta_{\rm 0f}$ is given by Eq.~\eqref{eq:theta_2f}. The figure
    shows a projection onto the $x$-$z$ plane. For a given $\theta_{\rm sd, i}$,
    the set of initial conditions forms a ring about CS2 (solid red area) and
    evolves to a ring about $\theta_{\rm 0f}$ (hatched red area) under
    nonadiabatic evolution. The final obliquities are the projection of this
    final cone onto the $\uv{l}$ axis (horizontal green
    lines).}}\label{fig:2nonadrot}
\end{figure}
\textcolor{Corr}{
When $\theta_{\rm sd, i}\neq 0$, the calculation of final obliquities is best
described by Fig.~\ref{fig:2nonadrot}, and yields the range of obliquities
\begin{equation}
    \abs{\theta_{\rm 0f} - \theta_{\rm sd, i}}
        \lesssim \sin\theta_{\rm f}
        \lesssim \theta_{\rm 0f} + \theta_{\rm sd, i}
    .\label{eq:nonad_q_f_dist}
\end{equation}
The absolute value in the lower bound is necessary to capture the behavior of
the projected obliquities in Fig.~\ref{fig:2nonadrot} when the hatched area
encloses $\uv{l}$.}

\textcolor{Corr}{
The process depicted in Fig.~\ref{fig:2nonadrot} has a physically clear
interpretation as a two-step process: (i) when $\eta$ is sufficiently large, the
evolution of the system is always adiabatic [$\dot{\theta}_2 \to 0$,
Eq.~\eqref{eq:dq2dt}], so $\uv{s}$ precesses around CS2 with constant angle
$\theta_{\rm sd, i}$ as $\eta$ decreases; (ii) when $\eta$ has decreased
sufficiently that the system becomes nonadiabatic [violating
Eq.~\eqref{eq:ad_base}], the obliquity of the system at that instant is frozen
in as the final obliquity.}

Fig.~\ref{fig:nonad_3_ensemble} shows the numerical result of $\theta_f$ vs
$\theta_{\rm sd,i}$ for $I =5^\circ$ and $\epsilon = 0.3$. We see that
Eq.~\eqref{eq:nonad_q_f_dist} provides good lower and upper bounds of the final
obliquity for $\theta_{\rm sd,i} \lesssim 45^\circ$.
